---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

Load the activity data from direct from the zip file provided for the assignment.
```{r, echo=FALSE}
 setwd("~/Documents/Data_Sci_Course/RepoducableResearch/RR_assessment1")
```

```{r load-data, echo=TRUE}
data <- read.csv(unz("activity.zip", "activity.csv"))
# convert date from factor to POSIXlt
data$date <- as.POSIXct(strptime(data$date, format="%Y-%m-%d"))
```

## What is mean total number of steps taken per day?

With the help of the dplyr package - the following calculates the mean, median, and sum of the steps recorded per day.

As stated in assignment - null values can be ignored for now - so I have removed them.

```{r , echo=TRUE, warning=FALSE, message=FALSE}
library(dplyr)

data.smy1 <- tbl_df(data) %>% 
                group_by(date) %>%
                summarise_at("steps", funs(sum(.,na.rm=TRUE)))

```

Lets take a look at the distribution of the total number of steps taken each day (sum).

```{r}
with(data.smy1, 
     hist(steps, 
          main="Histogram of total number \n of steps per day", 
          xlab="Total number of steps per day",
          ylab="Frequency of days")
     )
```

and lets get the median and mean of the total number of steps taken per day

```{r, echo=TRUE}
data.smy1.mean <- mean(data.smy1$steps)
data.smy1.median <- median(data.smy1$steps)
```

So the mean is [`r data.smy1.mean`] and the median is [`r data.smy1.median`]. 

## What is the average daily activity pattern?

Ok lets get the data into shape for plotting interval against steps. 

Using the dplyr package I will group by interval and take the mean number of steps recorded for that interval across the all the readings (from `r max(data$date)` to `r min(data$date)`).

```{r, echo=TRUE}
data.smy2 <- tbl_df(data) %>% 
                group_by(interval) %>% 
                summarise_at("steps", funs(mean(.,na.rm=TRUE)))
```

The data above is plotted onto the figure below.

```{r, echo=TRUE}
with(data.smy2, 
     {
             plot(x=interval,
                  y=steps, 
                  type="l",
                  col="blue", 
                  xlab="Interval (mins)",
                  ylab="Number of Steps",
                  main="Mean number of steps recorded per 5-min interval"
             )
     })
```

```{r, echo=TRUE}
data.smy2.maxline <- data.smy2[order(data.smy2$steps, decreasing=TRUE)[1],]
```

As can be seen on the chart the interval with the maximum mean number of steps taken is [`r data.smy2.maxline$steps`] from interval [`r data.smy2.maxline$interval`].

## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
